\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-04 paper/paper.tex
first saaturdy edits

Our research design is twofold.  First, we present visualizations that
describe various features of the Hahner twin's results, and an
important element in our visualizations is the difference between
runners' Rio times and their prior personal best marathon times.  Our
visualizations suggest that the Hahner twin's pace in the marathon was
slow, albeit not excessively so, but that their simultaneous finish
was quite unusual given the twins' differences in abilities (and
similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, which we develop based on relationships
across the Olympic marathoners between personal best times and
observed Rio finishing times.  Our simulations reinforce what we
observed in visualization, namely, that the Hahner twins did not run
appreciably slowly yet finished suspiciously close to each other.  We
return in our conclusion to Kurschilgen's claims about the Hahners and
offer thoughts about their validity.

@@

One way that we might assess whether an athlete's Rio finish was
unusual---or, say, whether two athletes' finishes were jointly
unusual---is by comparing observed finishing times on August 14 to a
measure of underlying marathon talent.  For the latter we use prior
personal best marathon times.  This is a natural measure of marathon
talent, we believe, but it is not entirely free of complications.  One
limitation of using personal best times as indicators of underlying
marathon skill is that these times are potentially confounded by the
marathon courses at which they were set; some courses, like the Berlin
marathon, are known for relatively fast times.  In addition, personal
best times may be confounded by conditions, like weather, that varied
across races.  Finally, personal best times may not capture Olympic
raceday idiosyncrasies that could affect individual runners.  For
example, a runner could have woken up with a minor cold on the morning
of August 14, 2016.  With these concerns in mind, we use half marathon
split times from the Olympic marathon as a secondary measure of
athletic skill in marathoning.  Our two measures of marathon talent
are featured in the visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and as function of their half
marathon split times.  The points in the plots are colored by
twin/triplet status, and both plots contain second-order polynomial
regression lines.  The ticks along the x-axis in both plots indicate times (either
personal best or half marathon split) of runners who earned a DNF
result.

@@

With respect to our second measure of marathon skill, Figure
\ref{fig:45degreeplot_half} shows that there was nothing abnormal
about the Hahner twins' overall finishing times, conditional on their
half marathon split times. As one might expect, a runner's time halfway through the race is a
fairly strong predictor of their finishing time at the end of the race.  Here we see that the relationship
between the twins' halfway time and their finishing time is similar to that
of the other runners.  The two points representing the Hahners are
in the middle of the distribution of points and therefore imply that the twins do not seem to have slowed down in any unusual way. 
 This is consistent with the previous
Figure \ref{fig:45degreeplot} and inconsistent with accusations made
against the Hahner twins, at least the part of the accusation that
focused on their overall pace.

Figure \ref{fig:residuals} gives a visualization of how unusual, or not,
the Hahner and Kim twins' finishing times were in the Olympic
marathon, conditional on our two measures of marathon talent.  The
left and right panels of the figure display the studentized residuals derived from the polynomial
regressions in Figure \ref{fig:scatter}'s panels, respectively.  Hence, we
compute the residuals from Figures \ref{fig:45degreeplot} and
\ref{fig:45degreeplot_half}, Studentize them, and arrange the result from least to most
along the x-axis in Figure \ref{fig:residuals} to display their cumulative distribution.

@@

The cumulative residuals depicted in Figure \ref{fig:residuals} are
consistent with our interpretation of the Hahner twin's Olympic
finishing times as not particularly remarkable.  With personal best
time as a measure of marathon talent, the two Hahner residuals are in
the right tail of the residual distribution; however, their locations
are not extreme.  One residual is located at the 88th percentile and
the other at the 93rd.  While they are in the tail end of the distribution, they are not the
major outliers that would lead us to reject the twins' claim.  
Moreover, the residuals for the Kim twins are similarly not
remarkable, although they are in the left tail of the residual
distribution, indicating that the Kims ran faster than one would have
expected.  If one uses half marathon split times as a measure of
marathon talent, similar results obtain.  Neither Hahner twin had a
finishing time that was particularly unusual given her half marathon
split, and this applies to the Kim twins as well.

@@

Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
pairs of runners.  Of these and ignoring the Hahner twins, ten had
exactly a 115 second gap in personal best times.  Of these ten pairs,
differences in finishing times, in seconds, are as follows: 36, 93, 172,
319, 379, 459, 552, 671, 675, and 739.  In other words, of all pairs
of runners in the Rio marathon who had a personal best difference that
was equivalent to the Hahner twins' difference, the twins had the
greatest compression based on finishing time.

@@

To estimate the conditional distribution of each Rio final results, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of marathon skill $X_i$. Here
we start by assuming that marathon skill can be measured by personal
best marathon time:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.  We
estimate $\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least
squares and all finishing Rio marathoners except for the Hahner/Kim
twins and Luik triplets.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anne and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her best time squared.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins}
  \label{fig:simdiff}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on half marathon split}
  \label{fig:simdiffhalf} 
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_half.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

>>>>

Our research design is twofold.  First, we present visualizations that
describe various features of the Hahner twin's results, and an
important element in our visualizations is the difference between
runners' Rio times and their prior personal best marathon times.  Our
visualizations suggest that the Hahner twin's pace in the marathon was
slow, albeit not excessively so, but that their simultaneous finish
was quite unusual given the twins' differences in abilities (and
similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, and our simulations reinforce what we
observed in prior visualizations, namely, that the Hahner twins did
not run appreciably slowly yet finished suspiciously close to each
other.  We return in our conclusion to Kurschilgen's claims about the
Hahners and offer thoughts about their validity.

@@

One way that we might assess whether an athlete's Rio finish was
unusual---or, say, whether two athletes' finishes were jointly
unusual---is by comparing a runner's observed Olympic women's marathon
finishing time on August 14, 2016, to a measure of her underlying
marathon talent.  For the latter we use prior personal best marathon
times.  This is a natural measure of marathon talent, we believe, but
it is not entirely free of complications.  For example, personal best
times are potentially confounded by the marathon courses on which they
were set; some courses, like the Berlin marathon, are known for
relatively fast times.  In addition, personal best times may be
confounded by conditions, like weather, that varied across races.
Finally, personal best times may not capture Olympic raceday
idiosyncrasies that could affect individual runners.  For example, a
runner could have woken up with a minor cold on the morning of August
14, 2016.  With these concerns in mind, we use a runner's half
marathon split time from the Olympic marathon as a secondary measure
of athletic skill in marathoning.  Our two measures of marathon talent
are featured in the visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain second-order polynomial regression lines.  The ticks along the
x-axis in both plots indicate times (either personal best or half
marathon split) of runners who earned  DNF results.

@@

With respect to our second measure of marathon skill, Figure
\ref{fig:45degreeplot_half} shows that there was nothing abnormal
about the Hahner twins' overall finishing times, conditional on their
half marathon splits. As one might expect, a runner's time halfway
through the course is a fairly strong predictor of her finishing time.
Here we see that the relationship between the Hehner twin's half
marathon split times and their finishing times is similar to that of
other Rio runners.  The two points representing the Hahners are in the
middle of the distribution of points and therefore imply that the
twins do not seem to have slowed down in any unusual way.  This is
consistent with the previous Figure \ref{fig:45degreeplot} and
inconsistent with accusations made against the Hahner twins, at least
the part of the accusation that focused on their overall pace.

Another perspective on the extent to which the Hahners' overall
marathon finishing times were not unusual can be gleaned from Figure
\ref{fig:residuals}.  This figure presents cumulative distributions of
the polynomial regression residuals from the regression models
displayed in the two panels of Figure \ref{fig:scatter}.  In
particular, we calculated the residuals from Figures
\ref{fig:45degreeplot} and \ref{fig:45degreeplot_half}, Studentized
them, and then arranged resulting Studentized residuals from least to
most along the horizontal axes in Figures
\ref{fig:studentizedresiduals} and \ref{fig:studentizedresiduals_half}.

@@

The cumulative residuals depicted in the two panels of Figure
\ref{fig:residuals} are consistent with our interpretation of the
Hahner twin's Olympic finishing times as not particularly remarkable.
With personal best time as a measure of marathon talent as in Figure
\ref{fig:studentizedresiduals}, the two Hahner residuals are in the
right tail of the residual distribution; however, their locations are
not extreme.  One residual is located at the 88th percentile and the
other at the 93rd.  While these two residuals are in the tail end of
the residual distribution, they are not the major outliers that would
lead us to think that the Hahner twin's marathon finishing times were
extremely slow.  Moreover, the residuals for the Kim twins are
similarly not remarkable, although they are in the left tail of the
residual distribution, indicating that the Kims ran faster than one
would have expected.  If one uses half marathon split times as a
measure of marathon talent, as in Figure
\ref{fig:studentizedresiduals_half}, similar results obtain.  Neither
Hahner twin had a finishing time that was particularly unusual given
her half marathon split, and this applies to the Kim twins as well.

@@

Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
pairs of runners.  Of these and ignoring the Hahner twins, ten had
exactly a 115 second gap in personal best times.  Differences in
finishing times of these ten pairs, in seconds, are as follows: 36,
93, 172, 319, 379, 459, 552, 671, 675, and 739.  In other words, of
all pairs of runners in the Rio marathon who had a personal best
difference that was equivalent to the Hahner twins' difference, the
twins had the greatest compression based on finishing time.

@@

To estimate the conditional distribution of each Rio final results, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of marathon skill $X_i$. Here
we start by assuming that marathon skill can be measured by personal
best marathon time:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.  We
estimate $\beta_1$, $\beta_2$, and $\sigma$ using ordinary least
squares and all finishing Rio marathoners except for the Hahner/Kim
twins and Luik triplets.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anne and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her best time squared.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on personal best times}
  \label{fig:simdiff}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on half marathon splits}
  \label{fig:simdiffhalf} 
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_half.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}
